An integrated distance sampling approach for modelling age-structured populations

### 2023-12-04 Manuscript_Nilsen_Nater.qmd
Updates to ms and .bib file

## Introduction

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005]. Over the last few decades, a tremendous progress have been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref). Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g. such as Essential Biodiversity Variables (ref).

Nevertheless, large scale and long term data sets are still relatively rare (ref), due to high costs of field data collection.

A cost efficient and well documented approach suitable is the use of line transect surveys. However, because the detectability is often less than 1 and variable between years and habitats,

-   Importance of abundance trends and demography

-   Short about methods -- detection probability -- marked vs unmarked animals

-   Approaches to estimate abundance and demography from unmarked animals

Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015]. These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e. the underlying vital rates. Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g. @schmidt2020].

In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds.

-   Estimate parameters with and without augmentation with demographic data -- compare estimates and precision
-   Estimate latent variable (juv summer survival) -- compare to literature values
-   Temporal variation in S and R -- via random effects models
-   Simulate - robustness

## Methods

### An integrated distance sampling model

@@

#### Age-structured population model

@@

#### Likelihoods for distance sampling data

@@

#### Likelihood for radio-telemetry data

@@

#### Hierarchical models with time-variation in parameters

@@

### Model testing with simulated data

@@

### Case study

@@

\[Figure 2: Map of Lierne study area?\]

### Bayesian implementation

@@

## Results

### Model performance on simulated datasets

@@

### Case study on willow ptarmigans in Lierne

@@

library(tidyverse)
library(tidybayes)

@@

test <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")

test1 <- test %>% spread_draws(Mu.R[])
mean_MuR <- mean(test1$Mu.R)
ci_MuR <- quantile(test1$Mu.R, probs = c(0.025, 0.975))

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021. The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at X (95% C.I = ), survival probability for February - July ($S_1$) was estimated as X (95% C.I = x), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at X. Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021. Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(mean_MuR, 1)` (95% C.I. = `r as.numeric(round(ci_MuR[1], 1))` - `r as.numeric(round(ci_MuR[2], 1))` ).

>>>>

## Abstract

*Estimation of abundance and demographic rates for populations of wild vertebrate species is a challenging but fundamental issue for both management and research into ecology and evolution. One approach that has been used extensively to estimate abundance of wild living populations is Distance Sampling (DS) methods based on line transect survey data. Historically, DS models were only available as open population models, and did not allow for the direct estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented. Here, we present an integrated distance sampling apporach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition. The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*). This framework allow us to make full use of the available data, and to effectively integrate the two data sources in an integrated modelling framework. Moreover, demographic rates often respond to environmental variation, and our approach allow us to directly estimate the effect of such environmental covariates on demographic rate variation. We first fit the model to simulated data, to assess it's ability to recover the underlying population dynamics when the true values are known. Then, we use data from a study of willow ptarmigan (Lagopus lagopus) in Norway as a case study.*

## 1. Introduction

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005]. Over the last few decades, a tremendous progress have been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref). Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g. such as Essential Biodiversity Variables [@Kissling2018].

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models (ref). While such methods can provide valuable information for both ecological research and management, they are also in many cases costly to implement over large areas. On the other hand, data from monitoring of abundance trends over time typically are based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for their popularity might be that they does not require repeated visits to the same sites to estimate detection probability (ref). Rather, it is assumed that detection probability decrease as the objects are located longer away from the transect line (ref). This makes them particularily useful for implementation in participatory monitoring program, allowing stakeholders to take part in the data collection process.

The most common implementation of the DS model has been to use close-population formulations, that does not allow to formulate a process model that project abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref). In recent years, DS approahces have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015]. Such approaches has greatly extended the application of DS approaches to study ecological dynamics in time and space. However, while these latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no stratight forward way to estimate demographic rates from such data. In cases where the age (and sex) of the observed individuals can be observed, this information can under some assumptions be used to inform about recruitment rates and survival probabilities (ref). For instance, in a integrated population model based on harvest statistics and observations of population age structure, both population abundance and demographic rates was estimated without even without additional data from marked individuals [@Nilsen2018].

Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g. @schmidt2020].

In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds. The process model is a stage-structured matrix model, that projects population size from one time step to the next based on the underlying survival and recruitment rates. This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates. In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities. The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref). In the paper, we first present the model and assess the robustness based on simulated data. In addition, we apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway. Because demographic rates are often affected by environmental covariates, we also illustrate how such covariates can be included in the modelling framework.

## 2. Methods

### 2.1 An integrated distance sampling model

@@

#### 2.1.1 Age-structured population model

@@

#### 2.1.2 Likelihoods for distance sampling data

@@

#### 2.1.3 Likelihood for radio-telemetry data

@@

#### 2.1.3 Hierarchical models with time-variation in parameters

@@

### 2.2 Model testing with simulated data

@@

### 2.3 Case study

@@

### 2.4 Bayesian implementation

@@

## 3. Results

### 3.1 Model performance on simulated datasets

@@

### 3.2 Case study on willow ptarmigans in Lierne

@@

library(tidyverse)
library(tidybayes)
source("tidySummary.R")

@@

out.data <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")

tidy_summary_stats <- tidySummary(out.data)

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021. The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper`), survival probability for February - July ($S_2$) was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper`), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S")$upper`). Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021. Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_R")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_R")$upper`).
